---
title: "VplotR advanced functions"
author: "Jacques Serizay"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VplotR advanced functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  eval = FALSE
)
```

## 0. Set up the session

```{r}
require(tidyverse)
require(VplotR)
require(ggplot2)
```

## 1. Import reads bam files and genomic loci files

We can import .bam files of interest from a given folder using the `importBamFiles()` function.

```{r eval = FALSE}
# Import bam file over annotated promoters, from a local folder. 
# WARNING: This takes quite a long time for libraries that are deeply sequenced
bam_files <- list.files('~/path/to/_bam-files/')
bam_list <- importPEBamFiles(
    bam_files, 
    shift_ATAC_fragments = TRUE
)
names(bam_list) <- bam_files
```

For this vignette, relevant data is also hosted on the Ahringer server:

```{r}
# Import ATAC-seq fragments 
bam_list <- readRDS(url('http://ahringerlab.com/VplotR/ATAC_PE_fragments.rds'))
lengths(bam_list)
```

We also need some relevant genomic locations. Let's get different sets of *C. elegans* promoters. 

```{r}
# Import annotated ce11 promoters 
ce_REs <- readRDS(url('http://ahringerlab.com/VplotR/ce11_annotated_REs.rds'))
proms_list <- readRDS(url('http://ahringerlab.com/VplotR/promoters_Granges.rds'))
lengths(proms_list)
```

## 2. Check distribution of fragment sizes over a specific set of segments

Let's have a look at the distribution of fragment sizes in each ATAC-seq library, over the different classes of promoters we have. For this, we can use the `getFragmentsDistribution()` function.

```{r}
# For each bam file, plot its distribution of fragment sizes over the tissue-specific REs or ubiquitous REs
sizes <- parallel::mclapply(names(bam_list), function(TISSUE) {
    getFragmentsDistribution(bam_list[[TISSUE]], proms_list, roll = 5)
}, mc.cores = length(bam_list)) %>%
    setNames(names(bam_list)) %>%
    namedListToLongFormat() %>% 
    setColNames(c('REs', 'Fragment_size', 'Number', 'Promoters', 'Sample')) %>%
    dplyr::mutate(Promoters = factor(Promoters, levels = names(proms_list))) %>% 
    dplyr::mutate(max_number = unlist(lapply(levels(Sample), function(TISSUE) {
        rep(max(Number[Sample == TISSUE]), sum(Sample == TISSUE))
    }))) %>% 
    dplyr::filter(Sample != 'mixed') %>% 
    dplyr::filter(Promoters == 'Ubiq._proms' | (gsub('_proms', '', .$Promoters) == .$Sample))
# Plot results
ggplot(sizes, aes(x = Fragment_size, y = Number, color = Promoters)) +
    geom_blank(aes(y = max_number)) + 
    geom_line() +
    theme_bw() +
    labs(
        title = 'Distribution of fragment sizes', 
        x = 'Fragment size',
        y = '# of fragments',
        color = 'Promoters classes'
    ) + 
    facet_wrap(Sample~Promoters, scales = "free") + 
    scale_color_manual(
        values = c('#991919', '#1232D9', '#3B9B46', '#D99B12', '#9e9e9e', '#D912D4')
    )
```

We can already notice that there is a lack of second/third/etc. nucleosome "bump" in somatic-specific ATAC-seq data over somatic REs. However, a more in-depth view of nucleosome binding would be desirable, to better understand the molecular context of promoter organization.

## 3. Vplots
### 3.1 Quick Vplot

Particularly, *where* fragments are mapping is an information as 
important as *how long* they are. The fragment length distribution plots generated
hereabove do not show the former parameter (where each fragment is mapping).
For this reason, we created a function to generate V-plots, 
to show distribution of ATAC fragments of variable sizes over a set of 
regions of interest. Such plot shows: 

- On the x axis: the position of the ATAC fragments (relative to the center of 
the loci of interest);
- On the y axis: the size of the ATAC fragments  

```r
Vmat <- bam_list[['Neurons']] %>% 
    computeVmat(proms_list[['Neurons_proms']]) %>%
    normalizeVmat(normFun = 'pctsum', roll = 3) %>% 
    plotVmat(
        main = 'Neurons-spe. ATAC fragments over Neurons-spe. proms'
    )
ggsave('examples/Vmat_Neurons-over-neurons-REs.pdf')
# Or do all at once in one function:
neurons_atac_neurons_proms <- plotVmat(
    bam_list[['Neurons']], 
    proms_list[['Neurons_proms']], 
    estimate_background = FALSE, 
    main = 'Neurons-spe. ATAC fragments over Neurons-spe. proms', 
)
gl_atac_gl_proms <- plotVmat(
    bam_list[['Germline']], 
    proms_list[['Germline_proms']], 
    estimate_background = FALSE, 
    main = 'Germline-spe. ATAC fragments over Germline-spe. proms', 
)
plots <- cowplot::plot_grid(gl_atac_gl_proms, neurons_atac_neurons_proms)
ggsave('examples/Germline_Neurons_Vmat-comparison.pdf', height = 6, width = 12, plots)
```

### 3.2 Normalized Vplots

Ideally, one would take into account the background ATAC-seq fragments to 
better estimate the molecular context in each condition.
Samples might have an background fragment size higher than others and this 
could affect the appearance of un-normalized Vplots.

```r
    list_params <- list(
        "Germline ATAC-seq over Ubiq. proms." = list(bam_list[['Germline']], proms_list[['Ubiq._proms']]),
        "Germline ATAC-seq over Germline proms." = list(bam_list[['Germline']], proms_list[['Germline_proms']]),
        "Neurons ATAC-seq over Ubiq. proms." = list(bam_list[['Neurons']], proms_list[['Ubiq._proms']]),
        "Neurons ATAC-seq over Neurons proms." = list(bam_list[['Neurons']], proms_list[['Neurons_proms']]),
        "Muscle ATAC-seq over Ubiq. proms." = list(bam_list[['Muscle']], proms_list[['Ubiq._proms']]),
        "Muscle ATAC-seq over Muscle proms." = list(bam_list[['Muscle']], proms_list[['Muscle_proms']]),
        "Hypod. ATAC-seq over Ubiq. proms." = list(bam_list[['Hypod.']], proms_list[['Ubiq._proms']]),
        "Hypod. ATAC-seq over Hypod. proms." = list(bam_list[['Hypod.']], proms_list[['Hypod._proms']]),
        "Intest. ATAC-seq over Ubiq. proms." = list(bam_list[['Intest.']], proms_list[['Ubiq._proms']]),
        "Intest. ATAC-seq over Intest. proms." = list(bam_list[['Intest.']], proms_list[['Intest._proms']])
    )
    parallel::mclapply(mc.cores = 5, c('libdepth', 'sum', 'max', 'quantile', 'Zscore'), function(FUN) {
        plots <- plotVmat(
            list_params, 
            xlim = c(-200, 200), 
            cores = length(list_params),
            normFun = FUN
        )
        p <- plots + 
            facet_wrap(~Cond., nrow = 2, dir = 'v') + 
            theme(legend.position = 'bottom') +
            theme(panel.spacing = unit(1, "lines"))
        ggsave(paste0('GL_Neurones_', FUN, '.pdf'), height = 7.2, width = 18)
    })
    #
    plots <- plotVmat(
        list_params, 
        xlim = c(-200, 200), 
        cores = length(list_params),
        normFun = 'libdepth', 
        universe = ce_REs
    )
    p <- plots + 
        facet_wrap(~Cond., nrow = 2, dir = 'v') + 
        theme(legend.position = 'bottom') +
        theme(panel.spacing = unit(1, "lines"))
    ggsave(paste0('GL_Neurones_libdepth_over-all-REs.pdf'), height = 7.2, width = 18)
    plots <- plotVmat(
        list_params, 
        xlim = c(-200, 200), 
        cores = length(list_params),
        normFun = 'libdepth', 
        universe = ce_REs[ce_REs$is.prom]
    )
    p <- plots + 
        facet_wrap(~Cond., nrow = 2, dir = 'v') + 
        theme(legend.position = 'bottom') +
        theme(panel.spacing = unit(1, "lines"))
    ggsave(paste0('GL_Neurones_libdepth_over-all-proms.pdf'), height = 7.2, width = 18)
```

To compute many different Vplots simultaneously, one can pass the two main 
arguments (bam_granges and granges) to the plotVmat function using a named list. 

```r 
list_params <- list(
    "Germline ATAC-seq over Ubiq. proms." = list(bam_list[['Germline']], proms_list[['Ubiq._proms']]),
    "Germline ATAC-seq over Germline proms." = list(bam_list[['Germline']], proms_list[['Germline_proms']]),
    "Neurons ATAC-seq over Ubiq. proms." = list(bam_list[['Neurons']], proms_list[['Ubiq._proms']]),
    "Neurons ATAC-seq over Neurons proms." = list(bam_list[['Neurons']], proms_list[['Neurons_proms']]),
    "Muscle ATAC-seq over Ubiq. proms." = list(bam_list[['Muscle']], proms_list[['Ubiq._proms']]),
    "Muscle ATAC-seq over Muscle proms." = list(bam_list[['Muscle']], proms_list[['Muscle_proms']]),
    "Hypod. ATAC-seq over Ubiq. proms." = list(bam_list[['Hypod.']], proms_list[['Ubiq._proms']]),
    "Hypod. ATAC-seq over Hypod. proms." = list(bam_list[['Hypod.']], proms_list[['Hypod._proms']]),
    "Intest. ATAC-seq over Ubiq. proms." = list(bam_list[['Intest.']], proms_list[['Ubiq._proms']]),
    "Intest. ATAC-seq over Intest. proms." = list(bam_list[['Intest.']], proms_list[['Intest._proms']])
)
plots <- plotVmat(
    list_params, 
    xlim = c(-200, 200), 
    cores = length(list_params),
    normFun = 'sum'
)
p <- plots + 
    facet_wrap(~Cond., nrow = 2, dir = 'v') + 
    theme(legend.position = 'bottom') +
    theme(panel.spacing = unit(1, "lines"))
ggsave('examples/Comparison_tissue-specific-normalized-Vmats.pdf', height = 7.2, width = 18)
```

## 4 Real case uses
### 4.1. Real case use #1: Compare nucleosome patterns over tissue-specific promoters or enhancers

Let's have a look at tissue-specific classes of promoters and the associated
Vplots, in a "programmatic" approach. 
We already presented the possibility to pass a list of arguments to 
plotVmat (see section 2.2). This allows a person to plot multiple Vplots at
once. Another way is to rely on parallelized computations of individual Vmats. 
Such simpler approach generating a list of plots allows the user to 
personalize each plot individually. 

```r
plotlist <- parallel::mclapply(seq_along(proms_list), function(K) {
    plot.proms <- parallel::mclapply(seq_along(bam_list), function(B) {
        message(names(bam_list)[B])
        plotVmat(
            bam_list[[B]], 
            proms_list[[K]],
            estimate_background = TRUE,
            main = paste0(names(bam_list)[B], '-spe. ATAC-seq over ', names(proms_list)[K], ' proms'), 
            breaks = seq(0, 12, length.out = 61), 
            ylim = c(55, 300)
        )
    }, mc.cores = length(bam_list))
    return(plot.proms)
}, mc.cores = length(proms_list))
plots <- purrr::flatten(plotlist)
p <- ggpubr::ggarrange(
    plotlist = plots, 
    nrow = 6, ncol = 6, 
    common.legend = TRUE,
    legend = "bottom"
)
ggsave('examples/Comparison-tissue-specific-normalized-Vmats_all-combinations.pdf', height = 45, width = 40)
```

It is striking that Germline promoters are characterized by prominent flanking 
nucleosomes, while somatic promoters are not. Ubiquitous promoters, 
on the other hand, are flanked by nucleosomes in all tissues. 

### 4.2. Real case use #2: Nucleosome enrichment estimation

We can estimate nucleosome enrichment using the `nucleosomeEnrichment()` function.
This function is used to quantify the local enrichment of a nucleosome at a 
certain distance from the midpoint of the genomic ranges of interest.

```r
list_scores <- parallel::mclapply(
    c("Germline", "Neurons", "Muscle", "Hypod.", "Intest."), 
    function(TISSUE) {
        message('>> ', TISSUE)
        nucenrich_tissue_spe_proms <- nucleosomeEnrichment(
            bam_granges = bam_list[[TISSUE]], 
            granges = proms_list[[paste0(TISSUE, '_proms')]], 
            estimate_background = TRUE, 
            verbose = TRUE
        )
        nucenrich_ubiq_proms <- nucleosomeEnrichment(
            bam_granges = bam_list[[TISSUE]], 
            granges = proms_list[['Ubiq._proms']], 
            estimate_background = TRUE, 
            verbose = TRUE
        )
        return(list(
            'tissue-spe-proms' = nucenrich_tissue_spe_proms, 
            'ubiq-proms' = nucenrich_ubiq_proms
        ))
    }, 
    mc.cores = 5
) %>% setNames(c("Germline", "Neurons", "Muscle", "Hypod.", "Intest."))
nucenrich_scores <- data.frame(
    tissue = factor(rep(names(list_scores), each = 2), levels = names(list_scores)), 
    promoters = c(rbind(paste0(names(list_scores), ' promoters'), rep('Ubiq. promoters', 5))), 
    promoters2 = factor(rep(c(0.5, 0.8), 5)),
    score = unlist(lapply(list_scores, function(Vmat) c(Vmat[[1]]$fisher_test$estimate, Vmat[[2]]$fisher_test$estimate))), 
    is_ubiq = factor(rep(c('Over tissue-specific promoters', 'Over ubiquitous promoters'), 5))
)
p <- ggplot(nucenrich_scores, aes(
    x = tissue, 
    y = score, 
    fill = tissue, 
    group = is_ubiq
)) + 
    geom_col() + 
    facet_wrap(~is_ubiq, nrow = 2) + 
    coord_flip() + 
    labs(title = 'Flanking nucleosome enrichment score', y = 'Enrichment score', x = 'Tissue-specific data') + 
    theme_bw() + 
    theme(legend.position = 'none')
ggsave('examples/Comparison_tissue-specific-nucleosome-enrichment.pdf', height = 5, width = 5)
```

This confirms what the Vplots suggest: flanking nucleosomes are well enriched 
at ubiquitous promoters in all tissues. Among tissue-specific promoters, 
only germline promoters have a significant enrichment of flanking nucleosome. 

### 4.3. Real case use #3: Use MNase-seq data

We can also use MNase data (only this data comes from mixed tissues) to look at nucleosome patterns. 

```r
# Read the bam files from local folder: 
### mnase_files <- list.files(
###     '~/20181127_PE-mnase-mapping/_bam-files', 
###     pattern = 'rep-1.*q10.bam', 
###     full.names = TRUE
### )
### MNase_mixed <- importPEBamFiles(
###     mnase_files, 
###     where = GenomicRanges::reduce(GenomicRanges::resize(ce_proms, width = 2000, fix = 'center')), 
###     shift_ATAC_fragments = FALSE
### ) %>% 
###     setNames(mnase_files %>% basename() %>% gsub('.map_pe\\^rm_chrM\\^rm_blacklist\\^q10.bam', '', .)) %>% 
###     '['(names(.) %>% gsub('YA_mnase_rep-1_|U', '', .) %>% as.numeric() %>% order())
# Or simply: 
MNase_mixed <- readRDS(
    url(
        'http://ahringerlab.com/VplotR/MNase_PE_fragments.rds'
    )
)
# Then plot Vmats
list_params <- lapply(seq_along(MNase_mixed), function(B) {
    lapply(seq_along(proms_list), function(K) {
        list(
            MNase_mixed[[B]], 
            resize(proms_list[[K]], 2000, fix = 'center')
        )
    }) %>% setNames(paste0('MNase (', gsub('YA_mnase_rep-1_', '', names(MNase_mixed))[B], ') coverage \nover ', names(proms_list), ' TSSs'))
}) %>% purrr::flatten()
plots <- plotVmat(
    list_params, 
    XAXIS.CUTOFF = 2000, 
    xlim = c(-1000, 1000), 
    ylim = c(75, 200),
    breaks = seq(0, 15, length.out = 60), 
    cores = length(list_params)
)
p <- plots + 
    facet_wrap(~Cond., ncol = length(proms_list), dir = 'h') + 
    theme(legend.position = 'bottom') +
    theme(panel.spacing = unit(1, "lines")) + 
    coord_fixed(ratio = 7)
ggsave('examples/MNase-coverage_proms.pdf', height = 15, width = 18)
```

As expected, nucleosomal reads are mapped over flanking nucleosomes, specifically
around ubiquitous and germline promoters. Arrays of nucleosomes are readily 
detected at these promoters. 

### 4.4. Compare nucleosome positioning at germline promoters, in WT or gon1gem2 mutants (where the promoters are not open)

```{r}
    
    gon1gem2_frags <- importPEBamFiles(
        '~/20200410_gon1gem2_ATAC/_bam-files/MS43.map_pe^rm_chrM^rm_blacklist^q10.bam', 
        shift_ATAC_fragments = FALSE
    ) 
    list_params <- list(
        "Germline-proms_WT" = list(bam_list[['Germline']], proms_list[['Germline_proms']]),
        "Germline-proms_gon1gem2" = list(gon1gem2_frags, proms_list[['Germline_proms']])
    )
    #
    plots <- plotVmat(
        list_params,
        xlim = c(-200, 200), 
        cores = length(list_params),
        normFun = 'sum'
    )
    Vplots <- plots + 
        facet_wrap(~Cond., nrow = 1, dir = 'v') + 
        theme(legend.position = 'bottom') +
        theme(panel.spacing = unit(1, "lines"))
    ggsave('germline-proms_WT-vs-gon1gem2.pdf', height = 3.6, width = 6)
    #
    list_scores <- list(
        'WT' = nucenrich_tissue_spe_proms <- nucleosomeEnrichment(
            bam_granges = bam_list[['Germline']], 
            granges = proms_list[['Germline_proms']], 
            verbose = TRUE
        ),
        'gon1gem2' = nucenrich_ubiq_proms <- nucleosomeEnrichment(
            bam_granges = gon1gem2_frags, 
            granges = proms_list[['Germline_proms']], 
            verbose = TRUE
        )
    )
    nucenrich_scores <- data.frame(
        tissue = factor(rep(names(list_scores), each = 2), levels = names(list_scores)), 
        promoters = c('germline', 'germline'), 
        promoters2 = factor(1, 2)),
        score = unlist(lapply(list_scores, function(Vmat) Vmat$fisher_test$estimate))
    ) 
    p <- ggplot(nucenrich_scores, aes(
        x = tissue, 
        y = score, 
        fill = tissue, 
        group = is_ubiq
    )) + 
        geom_col() + 
        facet_wrap(~is_ubiq, nrow = 2) + 
        coord_flip() + 
        labs(title = 'Flanking nucleosome enrichment score', y = 'Enrichment score', x = 'Tissue-specific data') + 
        theme_bw() + 
        theme(legend.position = 'none')
    ggsave('examples/Comparison_tissue-specific-nucleosome-enrichment.pdf', height = 5, width = 5)

```

